---
title: "4. RFID Tutorial: Quantifying feeder visits"
author: "Gabrielle Davidson"
date: "2024-08-15"
output: html_document
---
## 4. Get set up

### Open R Studio

### Check R packages installed
```{r echo=T, results='hide', error=FALSE, warning=FALSE, message=FALSE}
#load the following packages
library(tidyverse)
library(rmarkdown)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(stringi) #may not need this
library(magrittr) #may not need this

```
### Download resources necessary for the workshop
Download the resources from **[my github page](https://github.com/DrGLDavidson/TOBEUPDATED)**

## 4.1 import your dataset to your environment from your working directory. 

```{r}

#clear the global environment so we don't have any conflicts with the next steps 

rm(list = ls(all.names = TRUE))

#choose the appropriate working directory

setwd("F:/RWorkspace/GitHub/RFID-workshop/data/outputFiles")

#call your most recent dataset 

df<-read.delim("Masterdf_noRepeats.txt", header=TRUE)
head(df)
names(df)

```

### 4.2 total number of visits per feeder, per day
```{r}
#a new dataframe that counts the total number of visits per feeder per individual, per date. 
individualVisits<-df%>%
  count(feeder, TagID_hex, date, sort = TRUE) 

#rename n column "visits"
names(individualVisits)[names(individualVisits) == "n"] <- "visits"  #rename columns
 names(individualVisits)
#visualise the data on a figure using ggplot 

individualVisits%>%
 ggplot(aes(x = TagID_hex, y = visits, color = feeder))+  #specify axis labels and groupings
  labs(x = "Bird ID", y = "Total visits") + #specify axis labels
  geom_point(aes(shape=feeder), size=2) + # specify shape and size of points for the variable feeder
  theme_bw() +   theme(axis.text.x = element_text(angle=90)) #removes the default gray background and rotates x labels. 
#Note that each data point is a unique date. Therefore this plot gives you information not only on the number of visits they make on a given day, but also over how many days. 
#Because of the outliers, it's difficult to resolve what is happening between 0 and 20 visits. 

#Restrict axis limits and minimise datapoint overlap

individualVisits%>%
 ggplot(aes(x = TagID_hex, y = visits, color = feeder))+  #specify axis labels and groupings
  labs(x = "Bird ID", y = "Total visits") + #specify axis labels
  geom_point(aes(shape=feeder), size=2) + # specify shape and size of points for the variable feeder
  theme_bw() +   theme(axis.text.x = element_text(angle=90))+
  geom_point() + #removes the default gray background and rotates x labels.
 ylim(0, 25) + #specify axis limits
    geom_point(position = position_jitter(width = 0.3)) #jitter the points along a horizontal axis to minimise overlap

#can also remove shape for visualization

individualVisits%>%
 ggplot(aes(x = TagID_hex, y = visits, color = feeder))+  #specify axis labels and groupings
  labs(x = "Bird ID", y = "Total visits") + #specify axis labels
  geom_point(size=1) + # specify size ONLY of points for the variable feeder
  theme_bw() +   theme(axis.text.x = element_text(angle=90))+
  geom_point() + #removes the default gray background and rotates x labels.
 ylim(0, 25) + #specify axis limits
    geom_point(position = position_jitter(width = 0.3))

#a new dataframe that counts the proportion an individual visits a feeder relative to the other feeders, per day. 
ProportionVisits<-individualVisits%>%
  group_by(TagID_hex, date)%>%
  arrange(TagID_hex)%>%
  mutate(freq = visits / sum(visits))%>%
  ungroup() 

```

### 4.3 proportion of visits per feeder, per day
```{r}

#a new dataframe that counts the proportion an individual visits a feeder relative to the other feeders, per day. 
ProportionVisits<-individualVisits%>%
  group_by(TagID_hex, date)%>%
  arrange(TagID_hex)%>%
  mutate(proportion = visits / sum(visits))%>%
  ungroup() 


#plot proportions. Better to use the original dataframe as the following code counts each line as a visit

ggplot(df,aes(x=factor(TagID_hex),fill=factor(feeder)))+ 
  geom_bar(position="fill")+
  geom_text(aes(label=..count..),stat='count',position=position_fill(vjust=0.5), angle=90, size=3) +
  #scale_x_discrete (labels = c("F01","F02","F03")) +
  theme_bw() +   theme(panel.grid.major = element_blank(),
                       panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), 
                       panel.border = element_rect(linetype = "solid", colour = "black", size=.8)) +
  #scale_fill_manual("Feeder", values = c("1" = "gray49", "0" = "gray84"))+
  theme(text=element_text(size=12,  family="serif"),
        axis.ticks = element_line(colour = "black", size = .6)) +
  labs(x = "Bird ID") +
  labs(y = "Proportion of visits") +
  theme_bw() +   theme(axis.text.x = element_text(angle=90))

```

### **4.4 EXERCISE** Visits
## 4.4.1 Create a dataframe of the total visits per individual across the whole of your experiment, irrespective of feeder.
(hint: group_by)

## 4.4.2 Create a new dataframe that excludes individuals that visited less than 5 times.
(hint: section 3.5)

## 4.4.3 rename the column of the number of visits to "totalVisits".

## 4.4.4 create a figure of your choice representing the number of visits from your new dataframe.

## 4.4.5 create a dataframe with a list of birds we consider participants and save it as a csv file.

### 4.5 EXERCISE - Time intervals 

Previously we created a column that indicated how many seconds had passed since an individual's previous visit per day. A similar approach can be taken to calculate intervals between visits between individuals (i.e. how long has passed since the previous birds' visit)

### THIS NEEDS EDITING 

- clear your global environment

- reload your database "filteredVisitsFeeders.txt"

- change the datetime column to be a POSIXct class 

- create a new column that calculates the time difference from the previous row of a dataframe grouped by date and feeder. Dont forget to use the argument arrange() so the timedate is sequential in your dataframe

- create a dataframe consisting of individuals that visited a feeder equal or less than 1 second after the previous visitor, perhaps this suggests they are more likely to displace other birds from the feeder. 

- create a dataframe of a list of unique individuals that have landed on a feeder within 1 second of the previous visitor

- create a dataframe that counts the number of times each individual has landed on a feeder within 1 second of the previous visitor


```{r}
#clear global environment
rm(list = ls(all.names = TRUE))

#upload file 

df<-read.table(file="filteredVisitsFeeders.txt", sep="\t", header=TRUE)
names(df)
# change datetime column to be POSIXct class
class(df$datetime)
df$datetime <- as.POSIXct(df$datetime)
class(df$datetime)

df1<-df %>%
  arrange(datetime)%>%
  group_by(date, feeder) %>%
  mutate(timeBetweenID = datetime - lag(datetime))%>%
  arrange(datetime)%>%
  ungroup()

df1$timeBetweenID <- as.numeric(as.character(df1$timeBetweenID)) 
class(df1$timeBetweenID)

df2<-df1%>%
  filter(timeBetweenID <=1)

uniqueDisplacers<-unique(df2$RFID)
uniqueDisplacers<-as.data.frame(uniqueDisplacers)

individualDisplacements<-df2%>%
  count(RFID, sort = TRUE) 

```

### 4.6 EXERCISE - Correct and incorrect visits

## 4.6.1 Create a column called "correctChoice" where visits to F01 contain the character "Y" and visits to the other incorrect feeders contain the character "N"
(hint: filter, then cbind)
# END OF 4. RFID Tutorial: Quantifying feeder visits 

```{r}
sessionInfo()
```

